---
title: "BIOL806_FinalProject_KY"
author: "Kayla Young"
date: "2025-12-04"
output: html_document
---

## Introduction

Dairy cows have a natural behavior to ruminate, which occurs
in all ruminant animal species since they are foregut
fermenters. Rumination is the process of regurgitating
undigested food, chewing for further mechanical digestion
and swallowed again to increase surface area for feed
particles, as well as aid in saliva production to buffer the
acidic rumen pH to ensure a stable environment for the
microbes that are responsible for the fermentation process
(Paudyal et al. 2018; Kaufman et al. 2018; Schirmann et al.
2013). Rumination is also known to be an indicator on
well-being and is widely used in research, especially during
the transition period which is the three weeks before and
three weeks after calving. This stage is the critically most
important in a dairy cow’s life and its success determines
how well they will perform in lactation (Calamari et al.
2014; Drackley 1999). Technology has been increasing in the
dairy industry that has helped researchers learn more how to
utilize rumination to predict dairy cattle health. There are
many forms of technology that can measure different metrics
like rumination time, activity time, body temperature, water
intake, etc. These technologies essentially act like an
apple watch, fit bit or oura ring for humans but instead,
for cows. They come in many different wearable forms from
collars, ear tags, wristbands and more increasingly popular,
bolus’s that reside inside the cow’s reticulum (Schirmann et
al. 2013; Kaufman et al. 2018; Soriani et al. 2012).

In recent decades, there has been an increase in the
utilization of this advanced technology to help improve
dairy cow health and well-being. Not only does it combat
efficiency for farmers, but it also has the capabilities to
aid in disease detection that is still currently being
explored. As stated previously, the transition period is
such a critical time, as there are many metabolic changes
happening, and if detrimental enough, it can negatively
impact dairy cow health (Drackley 1999). Which in turn, has
a negative effect on farmers by losing product, increase in
vet care and if worse enough, loss of the cow. Once a cow
undergoes parturition, or gives birth, there is an increased
demand for nutrients since most of the cow’s energy is used
to produce mass amounts of milk, this is called a negative
energy balance (NEB). Dairy cows must eat to meet their
energy requirements and if failure to adapt to NEB puts them
at a higher risk for heath disorders or metabolic diseases
in the first few weeks of lactation (McArt et al. 2012;
Tufarelli et al. 2024). Some of these health disorders
include hypocalcemia (milk fever), ketosis, rumen acidosis,
displaced abomasum, retained placenta, etc. When dairy cows
get sick or feel unwell, that will have a negative effect on
their rumination and dry matter intake (DMI), causing them
to eat less. In addition, most metabolic diseases, some cows
can show no symptoms and be considered subclinical, which
makes it harder to detect just from visual observation.
However, with advanced dairy technology, rumination can be
measured, and farmers and researchers can now utilize this
to monitor dairy health data to help detect
diseases/disorders earlier (Soriani et al. 2012).

### Objective

The objective of this study is to explore rumination time in
the last week of pregnancy and see how it impacts milk
production in the first 30 days into lactation (days in
milk, DIM).

## Methods

The data used in this study was collected from the Fairchild
Dairy Teaching and Research Center and from the automated
cow monitoring system, SmaXtec, for (n=34) Holstein dairy
cows. Bolus’s have been installed for majority lactating
cows and have been collecting rumination time (min/24hr),
water intake (l/24hr), body temperature (°C), activity,
calving indicator, and heat index. The metrics of interest
for this study involved milk yield data from 0-30 DIM and
rumination time every 10 mins from -14d pre-partum to 30d
post-partum.

### Data Cleaning and Organizing

A list of cows were obtained from the farm manager that have
calved since April 1st, 2025, and stored in an excel
spreadsheet which consisted of Cow ID, lactation number,
date of calving, 14d before calving date, as well as the
date of when the cow reached 30 DIM. From there, individual
cow data (rumination time (min/24hr)) from the SmaXtec
online database was exported from -14d pre-partum to 30d
post-partum and briefly scanned to ensure it consisted
appropriate days needed. Cows that did not meet the date
requirements, i.e. bolus was administered at a later time
and not yet calibrated, was not included in the study
(n=13). In a new excel sheet, consisted of all SmaXtec
imported data, and new columns was created to gather 2 hour,
6 hour, and 24 hours mean rumination time (RT) for each
individual cow on each day. Each cow’s spreadsheet was then
copied and pasted to create one long running spreadsheet for
all cows in this study (n=34) to then be imported to R
studio for further analysis.

As for milk yield, data was exported from the milk system
(DeLaval Delpro) at the UNH dairy computer, for each cow
(n=34). Each excel spreadsheet consisted of Cow ID, date,
days since last calving (DIM), and total daily milk yield
(lbs.). New columns of lactation number, weekly milk yield
(MY) and Week (1-5) was added. Similar to the RT data, each
cow’s spreadsheet was then copied and pasted to create one
long running spreadsheet to then be imported to R studio for
further analysis.

### Rumination time in last week of pregnancy

Each cow was placed into one of two groups, above or below,
based off of the their individual mean rumination time (RT),
accounting for parity (primiparous or multiparous), as
primiparous cows differ biologically from multiparous cows,
in the last week of pregnancy. To start, each individual
cows 24 hour RT was selected from -7d to -1d, to obtain the
individual mean RT in the last week of pregnancy. For
primiparous cows, (lactation = 1, n = 5), each individual
mean RT was then averaged together to obtain a group mean RT
to classify each primparous cow if their individual mean RT
in the last week of pregnancy was above or below the group
RT average. This same step was also completed for the
multiparous cows, (lactation = 2+, n = 29).

### Rumination time in the week before and week after calving

Both the above and below groups mean rumination time was
observed in the week before and week after calving, with
data focusing from day -7 to 7, with calving being day 0.
Starting with the below group, daily (24hr) rumination time
was averaged from each cow, for each day from -7 to 7. The
same step was repeated for the cows in the above group. This
data was then combined together to be constructed into a
time series plot.

### Milk yield in first 30 days in milk

Groups above and below were assigned to the milk yield data
to explore lactation success in the first 30 days in milk
(DIM). This data was used to create two time series plots,
one for the above group and one for the below group, with
week and weekly milk yield (mean milk yield for each week).
Similar to comparing mean rumination time between the above
and below groups, the same was done to visualize the
difference in milk production in the first 30 DIM. For the
below group, the mean daily milk yield was averaged from day
1 to day 30. The same was completed for the above group,
then both data was then combined together to be constructed
into a time series plot.

## Results

### Data Organizing and Cleaning

All codes to creating the tables, figures and data frames
can be found here
<https://github.com/1006kay/Biol_806_FinalProject.git> on
github.

```{r, message = FALSE, warning= FALSE, echo=FALSE}
#Load Packages 
library(tidyverse) 
library(readxl) 
library(purrr) 
library(ggplot2) 
library(dplyr) 
library(lubridate) 
library(kableExtra)

RT30d_data <- read_excel("~/Desktop/Master Program/Fairchild Data /RT30d_data.xlsx") MY30d_data1 <- read_excel("~/Desktop/Master Program/Fairchild Data /MY30d_data1.xlsx")
```

### Rumination time in the last week of pregnancy

To account for parirty, primiparous (1st lactation) and
multiparous (2+ lactation) cows were grouped together to
obtain mean rumination time for their last week of pregnancy
(LWP), which consisted of days -7 to -1, relative to calving
at day 0. For primiparous cows, the mean RT in the LWP was
489.99 min/24hr. For multiparous cows, the mean RT in the
LWP was 473.12 min/24hr. In table 1, it shows the cows that
were enrolled in the data collection process that had an
individual mean RT that was below parity (primi- or
multiparous) average. Also included in table 1, is the
lactation each cow is currently in. The sample size of cows
placed in the below group is n = 16. Table 2 includes the
same metrics listed in table 1, however this table shows the
cows that had a greater mean RT in the LWP than the parity
RT average. The sample size of cows placed in the above
group is n = 18.

\pagebreak

```{r, echo=FALSE, tab.cap="The distribution of cows that belong in the below group based off of their rumination time (RT) in their last week of pregnancy. The primiparous cows (n=5) were grouped together to find the primiparous mean RT (489.99 min/24hr) and the multiparous cows (n=29) were grouped together to find the multiparous mean RT (473.12 min/24hr)."}
#1. Finding individual RT (24hr) in last week of pregnancy
ind_average24rt <- RT30d_data %>%
  select(`Cow ID`, Day, `RT 24hr`) %>% 
  filter(Day %in% c(-7, -6, -5, -4, -3, -2, -1)) %>%
  distinct() %>% #removes identical rows
  group_by(`Cow ID`) %>%
  summarise(`meanRT-1wk` = mean(`RT 24hr`)) #gets avg(mean) RT in -7 to -1d 

#2. Finding mean RT based off parity (primi = 1, multi = 2<), then creating above or below groups 
MYlactation <- MY30d_data1 %>%
  select(`Cow ID`, Lactation)%>%
  distinct() #data frame creating cow id and lactation (parity)

#Primi
ind_average24rt.1 <- left_join(MYlactation, ind_average24rt, by = "Cow ID") #combines lactation status to 24 hr RT data by Cow ID
primi_ind_avg <- ind_average24rt.1 %>%
  filter(Lactation == 1) %>% #separating primi cows
  mutate(mean_grp1avg_rt24 = mean(`meanRT-1wk`), 
         comparison = case_when(`meanRT-1wk` > mean_grp1avg_rt24 ~ "above",
      `meanRT-1wk` < mean_grp1avg_rt24 ~ "below",
      TRUE ~ "equal")) #indicates if ind cow falls above or below grp avg for primiparous cows
#multi
multi_ind_avg <- ind_average24rt.1 %>%
  filter(Lactation > 1) %>% #separating mutli cows
  mutate(mean_grp1avg_rt24 = mean(`meanRT-1wk`), 
         comparison = case_when(`meanRT-1wk` > mean_grp1avg_rt24 ~ "above",
      `meanRT-1wk` < mean_grp1avg_rt24 ~ "below",
      TRUE ~ "equal")) #indicates if ind cow falls above or below grp avg for multiparous cows

#combining both primi + multi data frames
ind_average24rt.2 <- rbind(primi_ind_avg, multi_ind_avg)
grp_avg_final <- ind_average24rt.2 %>%
  mutate(parity = case_when(Lactation <= 1 ~ "primi", Lactation > 1 ~ "multi"))

#Cow ID and Parity data frame
cow_parity <- grp_avg_final %>%
  select(`Cow ID`, parity)


grp1below <- grp_avg_final %>%
  filter(comparison == "below")

colnames(grp1below)[3] <- c("Mean Ind. RT in LWP")
colnames(grp1below)[4] <- c("Mean Parity RT in LWP")
colnames(grp1below)[5] <- c("Group")
colnames(grp1below)[6] <- c("Parity")

kable(grp1below)
```

\pagebreak

```{r, echo=FALSE, tab.cap="The distribution of cows that belong in the above group based off of their rumination time (RT) in their last week of pregnancy. The primiparous cows (n=5) were grouped together to find the primiparous mean RT (489.99 min/24hr) and the multiparous cows (n=29) were grouped together to find the multiparous mean RT (473.12 min/24hr)."}
grp1above <- grp_avg_final %>%
  filter(comparison == "above")

colnames(grp1above)[3] <- c("Mean Ind. RT in LWP")
colnames(grp1above)[4] <- c("Mean Parity RT in LWP")
colnames(grp1above)[5] <- c("Group")
colnames(grp1above)[6] <- c("Parity")

kable(grp1above)
```

\pagebreak

In Figure 1, shows the distribution of cows individual
rumination time after assigning groups, above or below,
based off of the mean RT in the last week of pregnancy.
Based off of Figure 1, the above group shows that there are
no outliers, the overall shape has some symmetric
distribution of individual mean RT since the median line is
centered while the whiskers are not evenly distributed, and
has a higher median when compared to the below group. As for
the below group, it is showing two outliers below the median
of the group, as well as the shape of this group is not
symmetric as the median is not centered but the whiskers are
similar in length.

```{r, echo=FALSE, fig.cap="The box plot figure shows the individual cow mean rumination time (min/24hr) in the last week of pregnancy (LWP) for both the above and below groups."}
above_below_bp <- ggplot(grp_avg_final, aes(x = comparison, y = `meanRT-1wk`, 
                             colour = comparison)) +
  geom_boxplot() +
  theme_light() +
  scale_color_manual(values = c("cadetblue3", "coral1")) +
  labs(x = "Group", 
       y = "Individual Cow Mean Rumination Time 
       (min/24hr) in the LWP",
       color = NULL) +
  theme(axis.title.y = element_text(size = 12),
        axis.title.x = element_text(size = 12))

print(above_below_bp)
```

\pagebreak

### Rumination time in the week before and week after calving

In Figure 2, it shows mean rumination time (min/24hr) by day
relative to calving. The below group has decreased mean
rumination time from day -7 to day 1, when compared to the
above group. After calving at day 0, the below group still
has a slight decrease in mean rumination time is comparison
to the above group. However, at day 2, both the above and
below groups have a similar mean rumination time just above
425 min/24hr. Both groups have a similar pattern of
rumination time throughout the two weeks. Around calving,
rumination time will decrease, then should increase shortly
after.

```{r, echo=FALSE, fig.cap="Time series plot that is showing the mean rumination (RT) (min/24hr) from the above and below group from 7 days before and 7 days after relative to calving (day 0)."}
#0. New data frames were created with all RT data for above and below groups for future analysis
grp1above_dataset <- RT30d_data %>%
  filter(`Cow ID` %in% c(976, 989, 1047, 1083, 1103, 1107, 1141, 1148, 1189, 1192, 1200,
                         1203, 1205, 1206, 1207, 1209, 1252, 1253))
grp1below_dataset <- RT30d_data %>%
  filter(`Cow ID` %in% c(953, 963, 1048, 1065, 1143, 1150, 1157,
                         1191, 1193, 1195, 1199, 1208, 1215, 1245, 1246, 1250))


#1. Creating data frame of 24hr RT for above group from -7 to 7 days relative to calving
grp1above_RT_LWP <- grp1above_dataset %>%
  select(`Cow ID`, Day, `RT 24hr`) %>%
  filter(Day %in% c(-7, -6, -5, -4, -3, -2, -1, 0, 1, 2, 3, 4, 5, 6, 7)) %>%
  distinct()

#2. Finding the mean 24hr RT for each day from the above group
grp1above_RT_LWP.1 <- grp1above_RT_LWP %>%
  group_by(Day) %>%
  summarise(above = mean(`RT 24hr`))

#3. Creating data frame of 24hr RT for below group from -7 to 7 days relative to calving
grp1below_RT_LWP <- grp1below_dataset %>%
  select(`Cow ID`, Day, `RT 24hr`) %>%
  filter(Day %in% c(-7, -6, -5, -4, -3, -2, -1, 0, 1, 2, 3, 4, 5, 6, 7)) %>%
  distinct() 

#4. Finding the mean 24hr RT for each day from the below group
grp1below_RT_LWP.1 <- grp1below_RT_LWP %>%
  group_by(Day) %>%
  summarise(below = mean(`RT 24hr`))

#5. Combining both above and below groups data frames to create long data frame to graph in ggplot
grp1_a_b_LWP <- left_join(grp1above_RT_LWP.1, grp1below_RT_LWP.1, by = "Day")
grp1_a_b_LWP_long <- grp1_a_b_LWP %>%
  pivot_longer(cols = c(above, below), 
               names_to = "group", values_to = "mean_RT")
#need to pivot_longer data to plot both above and below groups

#6. Creating ggplot point graph to visualise mean RT from -7d to 7d for the above group and below group
mean_a_b_rt <- ggplot(grp1_a_b_LWP_long, aes(x = Day, y = mean_RT, color = group)) +
  geom_point() +
  geom_line(aes(group = group)) +
  scale_x_continuous(breaks = -7:7) + #need this for xaxis in days
  scale_color_manual(values = c("cadetblue3", "coral1")) +
  theme_light() +
  labs(y = "Mean Rumination Time (min/24hr)", color = NULL) +
  theme(axis.title.y = element_text(size = 12),
        axis.title.x = element_text(size = 12))

print(mean_a_b_rt)
```

\pagebreak

### Milk yield in first 30 days in milk

In Figure 3, it shows the weekly milk yield (lbs) for the
above group in the first five weeks in lactation. Based off
of the figure, all cows have a similar trend throughout the
five weeks, increasing as the weeks increase. Cow 989 has a
slight decrease at week three but increases after. Cow 976
and 1047 both have a decrease in weekly milk yield at week
five.

```{r, echo=FALSE, fig.cap="Time series plot that shows the mean weekly milk yield (lbs) for each individual cow in the above group in the first five weeks into lactation. Calving is at the beginning of week 1."}
#1. Assign groups (above or below) to MY data
cowid_group <- grp_avg_final %>%
  select(`Cow ID`, comparison, parity)
MY30d_group1 <- left_join(MY30d_data1, cowid_group, by = "Cow ID")
#assigning group to MY data

#2. Creating seperate data frames for above and below group
grp1MY_above <- MY30d_group1 %>%
  filter(comparison == "above") %>%
  mutate(`Cow ID` = as.factor(`Cow ID`)) #for ease of plotting in ggplot

grp1MY_below <- MY30d_group1 %>%
  filter(comparison == "below") %>%
  mutate(`Cow ID` = as.factor(`Cow ID`))

#3. Graphs in ggplot point for above and below groups to visualise weekly mean MY in the first 5 weeks of lactation
#above
above_my <- ggplot(grp1MY_above, aes(x = Week, 
                         y = `Weekly MY`, 
                         color = `Cow ID`)) +
  geom_point() +
  geom_line() +
  theme_light() +
  labs(y = "Weekly Milk Yield (lbs) for 
       the Above Group") +
  theme(axis.title.y = element_text(size = 12),
        axis.title.x = element_text(size = 12),
        legend.text = element_text(size = 7)) +
  guides(color=guide_legend(nrow = 10, byrow = FALSE, order = 1))


print(above_my)
```

\pagebreak

In Figure 4, it shows the weekly milk yield (lbs) for the
below group in the first five weeks in lactation. Based off
of the figure, there are a some cows that have a decrease in
weekly milk yield. Cows 1150 has a decline in weekly milk
yield at week three, that continues to decrease. Cow 1195
also has a slight decrease at week three but slightly
increases at week four and five. Cow 1048 has a decrease in
weekly milk yield at week two, but increases in the next
week. Cow 1191 and 953 both have a slight decrease in weekly
milk yield at week five.

```{r, echo=FALSE, fig.cap="Time series plot that shows the mean weekly milk yield (lbs) for each individual cow in the below group in the first five weeks into lactation. Calving is at the beginning of week 1."}
#below
below_my <- ggplot(grp1MY_below, aes(x = Week, 
                         y = `Weekly MY`, 
                         color = `Cow ID`)) +
  geom_point() +
  geom_line() +
  theme_light() +
  labs(y = "Weekly Milk Yield (lbs) for 
       the Below Group") +
  theme(axis.title.y = element_text(size = 12),
        axis.title.x = element_text(size = 12),
        legend.text = element_text(size = 7)) +
  guides(color=guide_legend(nrow = 10, byrow = FALSE, order = 1))

print(below_my)
```

\pagebreak

In Figure 5, it shows that the cows in the below group,
which had a lower rumination time in the last week of
pregnancy, had slightly lower mean milk yield in the first
30 days into lactation when compared to the cows in the
above group, which had a higher rumination time in the last
week of pregnancy.

```{r, echo=FALSE, fig.cap="Time series plot that shows the mean milk yield (lbs) for the above and below groups in the first 30 days in milk (DIM). Calving is at day 0."}
#4. Finding mean MY for the above and below group
#above
avgaboveMY <- grp1MY_above %>%
  select(`Cow ID`, `Days Since Last Calving`, `Total Yield`) %>%
  group_by(`Days Since Last Calving`) %>%
  summarise(above = mean(`Total Yield`))

#below
avgbelowMY <- grp1MY_below %>%
  select(`Cow ID`, `Days Since Last Calving`, `Total Yield`) %>%
  group_by(`Days Since Last Calving`) %>%
  summarise(below = mean(`Total Yield`))

#5. Combining above and below MY data to graph
MY_above_below <- left_join(avgaboveMY, avgbelowMY, by = "Days Since Last Calving") 
MY_above_below_long <- MY_above_below %>%
  pivot_longer(cols = c(above, below),
               names_to = "group", values_to = "mean_my")


#6. Creating ggplot point graph on above and below mean MY in the first 30 days in milk (DIM)
mean_my_a_b <- ggplot(MY_above_below_long, aes(x = `Days Since Last Calving`, 
                                y = mean_my, 
                                color = group)) +
  geom_point() +
  geom_line() +
  scale_color_manual(values = c("cadetblue3", "coral1")) + 
  theme_light() +
  labs(x = "Days in Milk", 
       y= "Mean Milk Yield (lbs)", 
       color = NULL) +
  theme(axis.title.y = element_text(size = 12),
        axis.title.x = element_text(size = 12))

print(mean_my_a_b)
```

## Discussion

**Rumination time in last week of pregnancy**

From table 1 and table 2, there is a slightly even
distribution of cows in the above and below group, above
with (n = 18) and below with (n = 16). Primiparous cows had
a greater mean rumination time when compared to multiparous
cows, which is unexpected since multiparous cows tend to
spend more time ruminating. However, since the sample size
for primiparous cows was only five in total, two of the five
primiparous cows were placed in the above group because of
their high rumination time compared to the other three,
which could explain why the primiparous mean rumination time
is greater. As for the multiparous cows, n = 13 were placed
in the below group while the remaining n = 16 placed in the
above group. As for Figure 1, there is a clear difference of
the above and below group in relation to their individual
mean RT in the last week of pregnancy.

**Rumination time in the week before and week after
calving**

The mean rumination time for the below group from -7 to day
0 in Figure 2 shows a huge gap representing decreased
rumination time when compared to the above group. However,
both have a similar trend of reduced RT from -1 to 1 d,
which provides support that calving is known to cause a
decrease in rumination time. Only on day 2, both groups have
a similar mean RT but overall even after calving, the below
group stays consistently under the above group.

**Milk yield in the first 30 days in milk**

Throughout Figures 3 and 4, it shows the relationship of
weekly milk yield from week one to five for both groups.
There is a similar trend with both groups, when the week
increases, so does weekly milk yield. This is realistic for
a dairy cow, as they will reach peak milk yield around 40 to
70 DIM. In the below group, there are some cows that had
greater reductions in weekly milk yield throughout the first
five weeks. However, in the above group, there are only a
couple of cows that had very little decreased weekly milk
yield. Lastly, the same concept can be applied to Figure 5,
as time increases (days) so does milk yield. Similar to
Figure 2, Figure 5 shows that the below group is
consistently under the above group, meaning that it can be
hypothesized that cows that have low rumination time in the
last week of pregnancy, will most likely make less milk in
the first 30 days in milk than cows that have high
rumination time in the last week of pregnancy.

## Conclusion

This data gives helpful insight on how rumination time in
the last week of pregnancy can impact milk production in the
first 30 days into lactation. With a decrease in rumination
time in the last week of pregnancy, it could possibly be the
reason for a decrease in milk yield. Because health events
were not provided in this study, it is unknown if common
health disorders like hypocalcemia, ketosis, retain
placenta, etc., caused the reduction of milk yield for the
cows in the below group. Because the transition period is a
critical time for dairy cows, decreased rumination time
increases the risk of health disorders that consequently
decreases milk yield at the start of lactation. This study
opens up new ideas to explore on how rumination time in the
last week of pregnancy impacts lactation success.

\pagebreak

## Works Cited

Calamari, L., N. Soriani, G. Panella, F. Petrera, A. Minuti,
and E. Trevisi. 2014. “Rumination Time around Calving: An
Early Signal to Detect Cows at Greater Risk of Disease.”
Journal of Dairy Science 97 (6): 3635–47.
<https://doi.org/10.3168/jds.2013-7709>.

Drackley, James K. 1999. “Biology of Dairy Cows During the
Transition Period: The Final Frontier?” Journal of Dairy
Science 82 (11): 2259–73.
[https://doi.org/10.3168/jds.S0022-0302(99)75474-3](https://doi.org/10.3168/jds.S0022-0302(99)75474-3){.uri}.

Kaufman, E.I., V.H. Asselstine, S.J. LeBlanc, T.F. Duffield,
and T.J. DeVries. 2018. “Association of Rumination Time and
Health Status with Milk Yield and Composition in
Early-Lactation Dairy Cows.” Journal of Dairy Science 101
(1): 462–71. <https://doi.org/10.3168/jds.2017-12909>.

Leblanc, Stephen. 2010. “Monitoring Metabolic Health of
Dairy Cattle in the Transition Period.” Journal of
Reproduction and Development 56 (S): S29–35.
<https://doi.org/10.1262/jrd.1056S29>.

McArt, J.A.A., D.V. Nydam, and G.R. Oetzel. 2012.
“Epidemiology of Subclinical Ketosis in Early Lactation
Dairy Cattle.” Journal of Dairy Science 95 (9): 5056–66.
<https://doi.org/10.3168/jds.2012-5443>.

Neves, R.C., B.M. Leno, K.D. Bach, and J.A.A. McArt. 2018.
“Epidemiology of Subclinical Hypocalcemia in Early-Lactation
Holstein Dairy Cows: The Temporal Associations of Plasma
Calcium Concentration in the First 4 Days in Milk with
Disease and Milk Production.” Journal of Dairy Science 101
(10): 9321–31. <https://doi.org/10.3168/jds.2018-14587>.

Paudyal, S., F.P. Maunsell, J.T. Richeson, C.A. Risco, D.A.
Donovan, and P.J. Pinedo. 2018. “Rumination Time and
Monitoring of Health Disorders during Early Lactation.”
Animal 12 (7): 1484–92.
<https://doi.org/10.1017/S1751731117002932>.

Schirmann, K., N. Chapinal, D.M. Weary, L. Vickers, and
M.A.G. Von Keyserlingk. 2013. “Short Communication:
Rumination and Feeding Behavior before and after Calving in
Dairy Cows.” Journal of Dairy Science 96 (11): 7088–92.
<https://doi.org/10.3168/jds.2013-7023>.

Soriani, N, E Trevisi, and L Calamari. 2012. Relationships
between Rumination Time, Metabolic Conditions, and Health
Status in Dairy Cows during the Transition Period.

Tufarelli, Vincenzo, Nikola Puvača, Dragan Glamočić,
Gianluca Pugliese, and Maria Antonietta Colonna. 2024. “The
Most Important Metabolic Diseases in Dairy Cattle during the
Transition Period.” Animals 14 (5): 816.
<https://doi.org/10.3390/ani14050816>.

### R Packages Citations

Garrett Grolemund, Hadley Wickham (2011). Dates and Times
Made Easy with lubridate. Journal of Statistical Software,
40(3), 1-25. URL <https://www.jstatsoft.org/v40/i03/>.

Wickham H, Averick M, Bryan J, Chang W, McGowan LD, François
R, Grolemund G, Hayes A, Henry L, Hester J, Kuhn M, Pedersen
TL, Miller E, Bache SM, Müller K, Ooms J, Robinson D, Seidel
DP, Spinu V, Takahashi K, Vaughan D, Wilke C, Woo K, Yutani
H (2019). “Welcome to the tidyverse.” Journal of Open Source
Software, 4(43), 1686. <doi:10.21105/joss.01686>
<https://doi.org/10.21105/joss.01686>.

Wickham H, Bryan J (2025). readxl: Read Excel Files. R
package version 1.4.5,
<https://CRAN.R-project.org/package=readxl>.

Wickham H, Henry L (2023). purrr: Functional Programming
Tools. R package version 1.0.2,
<https://CRAN.R-project.org/package=purrr>.

H. Wickham. ggplot2: Elegant Graphics for Data Analysis.
Springer-Verlag New York, 2016.

Wickham H, François R, Henry L, Müller K, Vaughan D (2023).
dplyr: A Grammar of Data Manipulation. R package version
1.1.4, <https://CRAN.R-project.org/package=dplyr>.

Zhu H (2024). kableExtra: Construct Complex Table with
'kable' and Pipe Syntax. R package version 1.4.0,
<https://CRAN.R-project.org/package=kableExtra>.
